Testing perceptual accuracy in a U.S. general population survey using stacked bar charts

### 2023-12-21 SDSS-2023/testing-charts-accuracy.qmd
quarto changes

When combining responses from different surveys, weights are rescaled with respect to the total population, so that their order after combining still reflects their relative importance. We *combine* (rather than cumulate) a set of $\ell$ surveys $S_1$, $S_2$, ... $S_\ell$ (with $\ell \ge 2$) as described in @omuircheartaighCombiningSamplesVs2002, by multiplying weights in $S_i$ by $\lambda_i$, for $1 \le i \le \ell$.
$\lambda_i \in [0,1]$ with $\sum_i \lambda_i = 1$ is given as 
\begin{equation}
\lambda_i = \frac{n_i/d_i}{\sum_{j=1}^{\ell}n_j/d_j},
\end{equation}
where $n_i$ is the nominal sample of survey $S_i$ and $d_i$ are the design effects for the estimators.
Here, $d_i$ are estimated as 
\begin{equation}
d_i = 1 + CV(w \in S_i)^2
\end{equation}
where $CV$ is the coefficient of variation of the weights $w$ within each sample, and is estimated as in @kishSurveySampling1965 : 
\begin{equation}
CV(w \in S) = \frac{\widehat{Var(w)}}{\bar{w}^2}.
\end{equation}

The data for this paper were collected in several rounds.
The resulting number of participants, effective sample sizes, and $\lambda$ values are shown in \autoref{tbl-rounds}.

| Name      | Date       |      \# Participants |                Effective Sample Size |         Sum of weights $\sum_i w_i$ |                                  $\lambda_i$ |
|------------|------------|-----------:|-----------:|-----------:|-----------:|
| Round 1   | April 2022 | `r nrow(round1_adj)` | `r round(neff(round1_adj$weight),1)` | `r round(sum(round1_adj$weight),1)` |        $\lambda_1$ = `r round(lambdas[1],3)` |
| Round 2   | May 2022   | `r nrow(round2_adj)` | `r round(neff(round2_adj$weight),1)` | `r round(sum(round2_adj$weight),1)` | $\lambda_2$ = `r sprintf("%.3f",lambdas[2])` |
| Round 3   | June 2022  | `r nrow(round3_adj)` | `r round(neff(round3_adj$weight),1)` | `r round(sum(round3_adj$weight),1)` |        $\lambda_3$ = `r round(lambdas[3],3)` |
| **Combined** | ---        |  `r nrow(round_123)` |  `r round(neff(round_123$weight),1)` |                                     |                                              |

: Survey rounds: dates, number of participants (nominal sample size), effective sample size, sum of weights, and factors for the combining of surveys.
{#tbl-rounds}

@@

A total of `r sum(c(nrow(round1_adj), nrow(round2_adj), nrow(round3_adj)))` respondents participated across the three rounds.
The number of responses and corresponding effective sample sizes in each round are shown in \autoref{tbl-rounds}.
All responses were combined into one set of survey responses, with adjusted combined sample weights and indicators for which task each respondent was exposed to.
The resulting sample sizes corresponding to each task are shown in \autoref{tbl-tasks}.

@@

gender <- 
  gender +
  annotate("text", x = 0.83, y = 5, label="ACS estimate & 95% C.I.", hjust=0, 
           size = 5, colour = "grey20") +
  annotate("segment", x = 0.5, xend=0.7, y = 5,yend=5, 
           linewidth = 1, colour="grey20") +
  annotate("segment", x = 0.5, xend=0.7, y = 5.1,yend=5.1, 
           linewidth = 1, colour="grey20") +
  annotate("segment", x = 0.5, xend=0.7, y = 4.9,yend=4.9, 
           linewidth = 1, colour="grey20") +
  annotate("text", x = 0.83, y = 15, label="Horizontal design", hjust=0,
           size = 5, colour = "grey20") +
  annotate("segment", x = 0.8333, xend = 0.8333, y = 17, yend=40, 
           linewidth = 0.75, colour=rounds_col[1], linetype=2) +
  annotate("text", x = 1, y = 20, label="Horizontal wide", hjust=0,
           size = 5, colour = "grey20") +
  annotate("segment", x = 1, xend = 1, y = 22, yend=40, 
           linewidth = 0.75, colour=rounds_col[2], linetype=2) +
  annotate("text", x = 1.17, y = 25, label="Vertical design", hjust=0,
           size = 5, colour = "grey20") +
  annotate("segment", x = 1.166, xend = 1.166, y = 27, yend=40, 
           linewidth = 0.75, colour=rounds_col[3], linetype=2)

@@

We first investigate respondent accuracy in selecting the correct response.
As participants were able to select the option 'They are the same', there are several ways to model accuracy and response.
The argument could be made that for the purposes of practical interpretation, 'they are the same' is a correct choice, as the options are visually very similar and not substantially different values within the context of the data shown in the chart.
However, we are interested in understanding whether viewers *can* perceive the difference and correctly identify which piece is larger and thus consider multiple ways of modeling response to investigate participant response patterns.
We begin by defining a measure of binary 'correctness', for which all answers that are not the correct option (B \[D\] is bigger) are 'incorrect', including the selection of 'they are the same'.

@@

gg2 <- round_123 %>%
  mutate(
    `Unaligned: A or B?` = ifelse(CorrectAB, "Correct", "Not correct"),
    `Aligned: C or D?` = ifelse(CorrectCD, "Correct", "Not correct")
    ) %>%
  pcp_select(`Unaligned: A or B?`, `Aligned: C or D?`) %>%
  pcp_scale() %>%
  pcp_arrange() %>%
  ggplot(aes_pcp()) +
    geom_pcp_axes() +
    geom_pcp(aes(colour = CorrectABCD), alpha = 0.2)+
    geom_pcp_boxes(fill=NA, colour = "black", linewidth=0.25) +
    geom_pcp_labels(colour = "black") +
  scale_y_continuous("Number of (effective) Respondents", breaks=seq(0,1,by=0.25),
                     labels = round(sum(round_123$weight)*seq(0,1,by=0.25),0),
                     sec.axis = sec_axis(~.*100,
                          name="Percent responses")) +
  xlab("") +
  scale_colour_manual(values = c("grey", "#84AC84", "#70A570", "forestgreen")) +
  ggtitle("(b) Combinations of responses") +
  theme(legend.position = "none")

@@

@fig-response-123 shows the results of a cell-means model with ordinal response $Y_k$, where $Y_k$ is the $k$th participant's response, $Y_k \in \{1, 2, 3\}$, where 'correct' is encoded as 1, 'they are the same' is encoded as 2, and 'wrong' is encoded as 3:

@@

where $\ell \in \{1, 2\}$; $i \in \{1, 2\}$ is the comparison type (1 = Aligned, 2 = Unaligned), and $j \in \{1, 2, 3\}$ is the chart design, with 1 = Vertical, 2 = Horizontal, and 3 = Horizontal wide.

@@

To formalize this pattern in a model, let $Y_{jk}$ describe the zooming behavior of panelist $k$ on task $j$.
We model zooming behavior (no = 0, yes = 1) as a logistic regression by correctness of response ($\rho$), task ($\tau$), and design ($\delta$) of the chart:

@@

Using linear scores for the response of certainty, with 'not certain at all' assigned a score of 1 and 'extremely certain' assigned a score of 5, we can estimate the effects of task, chart design, and correctness on certainty by using a cell-means model of the form: 
\begin{equation}
Y_{k} = \mu_{ij\ell(k)} + \epsilon_{k},
\end{equation}
where $k = 1, ..., N$, $\mu_{ij\ell(k)}$ is average certainty (measured on a scale from 1 to 5) of the four combinations of task and correctness by each of the three designs, where $i = 1, 2$ encodes unaligned/aligned, and $j=1,2$ encodes wrong, correct, and $\ell = 1, 2, 3$ encodes horizontal, wide horizontal, and vertical stacked bar charts, respectively.
We also assume that errors are normally distributed, i.e. $\epsilon_k \stackrel{i.i.d}{\sim} N(0, \sigma^2)$ for all $k = 1, ..., N$.
The results are shown in @fig-certainty.
We find that the highest scores for certainty are associated with the aligned task, but certainty scores are not significantly different between correct and wrong responses.
Scores are (mostly) significantly lower for the unaligned task.
Interestingly, the lowest certainty scores are associated with correct responses to the unaligned task; although respondents successfully complete the task, they are less certain in their selection.

@@

Let $Y_k$ be the response of participant $k$, on a scale from 1 = 'correct', 2 = 'they are the same' to 3 = 'wrong'. We use a generalized cumulative logistic regression, where $\mu_i$ are intercepts $1 \le i < 3$, $X_k$ are demographics of the $k$th participant (in form of the model matrix), and $\beta_{i}$ are the coefficients.

\begin{equation}
\text{ logit } P(Y_k \le i \mid X_k) = \mu_i + X_{k}'\beta_{i}
\end{equation}

>>>>

When combining responses from different surveys, weights are rescaled with respect to the total population, so that their order after combining still reflects their relative importance. We *combine* (rather than cumulate) a set of $\ell$ surveys $S_1$, $S_2$, ... $S_\ell$ (with $\ell \ge 2$) as described in @omuircheartaighCombiningSamplesVs2002, by multiplying weights in $S_i$ by $\lambda_i$, for $1 \le i \le \ell$.
$\lambda_i \in [0,1]$ with $\sum_i \lambda_i = 1$ is given as 
\begin{equation}
\lambda_i = \frac{n_i/d_i}{\sum_{j=1}^{\ell}n_j/d_j},
\end{equation}
where $n_i$ is the nominal sample of survey $S_i$ and $d_i$ are the design effects for the estimators.
Here, $d_i$ are estimated as 
\begin{equation}
d_i = 1 + CV(w \in S_i)^2
\end{equation}
where $CV$ is the coefficient of variation of the weights $w$ within each survey sample $S_i$, and is estimated as in @kishSurveySampling1965 : 
\begin{equation}
CV(w \in S) = \frac{\widehat{Var(w)}}{\bar{w}^2},
\end{equation}
and where $\bar{w}$ and $\widehat{Var(w)}$ are sample mean and sample variance of the weights $w$ of survey sample $S$, respectively.

The data for this paper were collected in several rounds.
The resulting number of panelists, effective sample sizes, and $\lambda$ values are shown in \autoref{tbl-rounds}.

| Name      | Date       |      \# Panelists |                Effective Sample Size |         Sum of weights $\sum_i w_i$ |                                  $\lambda_i$ |
|------------|------------|-----------:|-----------:|-----------:|-----------:|
| Round 1   | April 2022 | `r nrow(round1_adj)` | `r round(neff(round1_adj$weight),1)` | `r round(sum(round1_adj$weight),1)` |        $\lambda_1$ = `r round(lambdas[1],3)` |
| Round 2   | May 2022   | `r nrow(round2_adj)` | `r round(neff(round2_adj$weight),1)` | `r round(sum(round2_adj$weight),1)` | $\lambda_2$ = `r sprintf("%.3f",lambdas[2])` |
| Round 3   | June 2022  | `r nrow(round3_adj)` | `r round(neff(round3_adj$weight),1)` | `r round(sum(round3_adj$weight),1)` |        $\lambda_3$ = `r round(lambdas[3],3)` |
| **Combined** | ---        |  `r nrow(round_123)` |  `r round(neff(round_123$weight),1)` |                                     |                                              |

: Survey rounds: dates, number of panelists (nominal sample size), effective sample size, sum of weights, and factors for the combining of surveys.
{#tbl-rounds}

@@

A total of `r sum(c(nrow(round1_adj), nrow(round2_adj), nrow(round3_adj)))` panelists participated across the three rounds.
The number of responses and corresponding effective sample sizes in each round are shown in \autoref{tbl-rounds}.
All responses were combined into one set of survey responses, with adjusted combined sample weights and indicators for which task each respondent was exposed to.
The resulting sample sizes corresponding to each task are shown in \autoref{tbl-tasks}.

@@

gender <- 
  gender +
  annotate("text", x = 0.83, y = 5, label="ACS estimate & 95% C.I.", hjust=0, 
           size = 4.5, colour = "grey20") +
  annotate("segment", x = 0.5, xend=0.7, y = 5,yend=5, 
           linewidth = 1, colour="grey20") +
  annotate("segment", x = 0.5, xend=0.7, y = 5.1,yend=5.1, 
           linewidth = 1, colour="grey20") +
  annotate("segment", x = 0.5, xend=0.7, y = 4.9,yend=4.9, 
           linewidth = 1, colour="grey20") +
  annotate("text", x = 0.83, y = 15, label="Horizontal design", hjust=0,
           size = 4.5, colour = "grey20") +
  annotate("segment", x = 0.8333, xend = 0.8333, y = 17, yend=40, 
           linewidth = 0.75, colour=rounds_col[1], linetype=2) +
  annotate("text", x = 1, y = 20, label="Horizontal wide", hjust=0,
           size = 4.5, colour = "grey20") +
  annotate("segment", x = 1, xend = 1, y = 22, yend=40, 
           linewidth = 0.75, colour=rounds_col[2], linetype=2) +
  annotate("text", x = 1.17, y = 25, label="Vertical design", hjust=0,
           size = 4.5, colour = "grey20") +
  annotate("segment", x = 1.166, xend = 1.166, y = 27, yend=40, 
           linewidth = 0.75, colour=rounds_col[3], linetype=2)

@@

We first investigate respondent accuracy in selecting the correct response.
Participants were able to select the option 'They are the same', there are several ways to model accuracy and response.
For the purposes of practical interpretation, the argument could be made that 'they are the same' is a correct choice, as the options are visually very similar and not substantially different values within the context of the data shown in the chart.
However, we are interested in understanding whether viewers *can* perceive the difference and correctly identify which piece is larger and thus consider multiple ways of modeling response to investigate participant response patterns.
We begin by defining a measure of binary 'correctness', for which all answers that are not the correct option (B \[D\] is bigger) are 'incorrect', including the selection of 'they are the same'.

@@

gg2 <- round_123 %>%
  mutate(
    `Unaligned: A or B?` = ifelse(CorrectAB, "Correct", "Not correct"),
    `Aligned: C or D?` = ifelse(CorrectCD, "Correct", "Not correct")
    ) %>%
  pcp_select(`Unaligned: A or B?`, `Aligned: C or D?`) %>%
  pcp_scale() %>%
  pcp_arrange() %>%
  ggplot(aes_pcp()) +
    geom_pcp_axes() +
    geom_pcp(aes(colour = CorrectABCD), alpha = 0.2)+
    geom_pcp_boxes(fill=NA, colour = "black", linewidth=0.25) +
    geom_pcp_labels(colour = "black") +
  scale_y_continuous("Number of (effective) Respondents", breaks=seq(0,1,by=0.25),
                     labels = round(sum(round_123$weight)*seq(0,1,by=0.25),0),
                     sec.axis = sec_axis(~.*100,
                          name="Percent responses")) +
  xlab("") +
  scale_colour_manual(values = c("grey", "#84AC84", "#70A570", "forestgreen")) +
  ggtitle("Combinations of responses") +
  theme(legend.position = "none")

@@

@fig-response-123 shows the results of a cell-means model with ordinal response $Y_{ik}$, where $Y_{ik}$ is the $k$th response, $1 \le k \le N$ for $N$ number of (nominal) responses. 
We use an encoding of $Y_k \in \{1, 2, 3\}$ for 'correct'=1, 'they are the same'= 2, and 'wrong' = 3:

@@

where $1 \le \ell < 3$; $i \in \{1, 2\}$ is the comparison type (1 = Aligned, 2 = Unaligned), and $j \in \{1, 2, 3\}$ is the chart design, with 1 = Vertical, 2 = Horizontal, and 3 = Horizontal wide. Note that we are using the notation $ij\ell(k)$ to emphasize the functional relationship between the $k$th response and the model parameters while avoiding double-subscripts, i.e. $ij\ell(k)$ signifies the index combination of $i, j,$ and $\ell$ of the $k$th response, i.e. $ij\ell(k)$ is 123, if the $k$th response was a wrong response ($\ell$=3) for an aligned comparison ($i$=1) in a horizontal ($j$=2) barchart.

@@

To formalize this pattern in a model, let $Y_k$ describe the zooming behavior on the $k$th response.
We model zooming behavior (no = 0, yes = 1) as a logistic regression by correctness of response ($\rho$), task ($\tau$), and design ($\delta$) of the chart:

@@

Using linear scores for the response of certainty, with 'not certain at all' assigned a score of 1 and 'extremely certain' assigned a score of 5, we can estimate the effects of task, chart design, and correctness on certainty by using a cell-means model of the form: 
\begin{equation}
Y_{k} = \mu_{ij\ell(k)} + \epsilon_{k},
\end{equation}
where $1 \le k \le N$ for $N$ responses, $\mu_{ij\ell(k)}$ is average certainty (measured on a scale from 1 to 5) of the four combinations of task and correctness by each of the three designs, for task $i = 1, 2$ (unaligned/aligned), design $j=1, 2, 3$ encoded as horizontal, wide horizontal, and vertical stacked bar charts, and  participant's decision $\ell = 1, 2$, encoded as correct or wrong, respectively.
We also assume that errors are normally distributed, i.e. $\epsilon_k \stackrel{i.i.d}{\sim} N(0, \sigma^2)$ for all $1 \le k \le N$.
The results are shown in @fig-certainty.
We find that the highest scores for certainty are associated with the aligned task, but certainty scores are not significantly different between correct and wrong responses.
Scores are (mostly) significantly lower for the unaligned task.
Interestingly, the lowest certainty scores are associated with correct responses to the unaligned task; although respondents successfully complete the task, they are less certain in their selection.

@@

Let $Y_k$ be the response of participant $k$, on a scale from 1 = 'wrong', 2 = 'they are the same' to 3 = 'correct'. We use a generalized cumulative logistic regression, where $\mu_\ell$ are intercepts $1 \le \ell < 3$, $X_k$ are demographics of the $k$th participant (the $k$th row of the corresponding model matrix), and $\beta_{\ell}$ are the coefficients.

\begin{equation}
\text{ logit } P(Y_k \le \ell \mid X_k) = \mu_\ell + X_{k}\beta_{\ell}
\end{equation}
